### 2014-05-12 paper/main.tex
Add beginnings of ACL-style paper

%
% File acl2014.tex
%
% Contact: koller@ling.uni-potsdam.de, yusuke@nii.ac.jp
%%
%% Based on the style files for ACL-2013, which were, in turn,
%% Based on the style files for ACL-2012, which were, in turn,
%% based on the style files for ACL-2011, which were, in turn, 
%% based on the style files for ACL-2010, which were, in turn, 
%% based on the style files for ACL-IJCNLP-2009, which were, in turn,
%% based on the style files for EACL-2009 and IJCNLP-2008...

%% Based on the style files for EACL 2006 by 
%%e.agirre@ehu.es or Sergi.Balari@uab.es
%% and that of ACL 08 by Joakim Nivre and Noah Smith

\documentclass[11pt]{article}
\usepackage[utf8]{inputenc}
\usepackage{acl2014}
\usepackage{times}
\usepackage{microtype}
\usepackage{url}
\usepackage{latexsym}
\usepackage{amsmath}
\usepackage{paralist} % inline lists
\usepackage{commands}

% smart quotes
\usepackage[autostyle, english=american]{csquotes}
\MakeOuterQuote{"}

%\setlength\titlebox{5cm}

% You can expand the titlebox if you need extra space
% to show all the authors. Please do not make the titlebox
% smaller than 5cm (the original size); we will check this
% in the camera-ready version and ask you to change it back.

\title{Deep Neural Models for Bilingual Lexicon Extraction}

\author{Jon Gauthier \\
  Symbolic Systems Program \\
  Stanford University \\
  Stanford, CA 94305 \\
  {\tt jgauthie@stanford.edu} \\\And
  Arthur Tsang \\
  Department of Computer Science \\
  Stanford University \\
  Stanford, CA 94305 \\
  {\tt atsang2@stanford.edu} \\}

\date{}

\begin{document}
\maketitle

\begin{abstract}
  We present a new method for producing bilingual lexicons from very
  non-parallel corpora. In contrast with previous approaches, our method
  separates the steps of \begin{inparaenum}[(1)]
    \item constructing distributed word representations from monolingual
      corpora, and
    \item building a mapping between these representations.
  \end{inparaenum}

  % TODO
  We demonstrate that a deep neural model can effectively learn a translational
  mapping between two monolingual vector spaces for the purposes of bilingual
  lexicon extraction. We evaluate the performance of this model relative to X
  and Y.
\end{abstract}

\section{Introduction}

In the task of bilingual lexicon extraction (BLE), we accept as input some pair
of corpora in different languages---perhaps parallel or perhaps not---and learn
associations of translational equivalence between one language (a "source
language") and the other (a "target language").

Bilingual lexicon extraction is obviously of most utility when applied to
under-resourced language pairs which do not already benefit from a surplus of
lexicon materials. These same language pairs, lacking resources as simple as
translation dictionaries, also lack sufficient aligned text for training
BLE models. For this reason, most recent work on this task focuses on
minimally supervised approaches which do not require parallel corpora
\cite{rapp1995,peirsman2010}.

\newcite{rapp1995} suggested that there is "a correlation between the patterns
of word co-occurrences in different languages." In line with this hypothesis,
recent work has determined the likelihood of two words as translations of one
another by comparing the translations of the contexts in which they appear. The
standard approach is to manually\unclear construct a "bilingual vector space" which
contains distributed word representations of words in both the source and target
language \cite{fung1998,peirsman2010,vulic2013}.

\section*{Acknowledgments}

The acknowledgments should go immediately before the references.  Do
not number the acknowledgments section. Do not include this section
when submitting your paper for review.

\bibliographystyle{acl}
\bibliography{bibliography}

\end{document}

### 2014-05-13 tools/build_vsm.py
word2vec command line: use as many workers as there are CPU cores

@@ -1,5 +1,6 @@
 import argparse
 import logging
+import multiprocessing
 
 from gensim.corpora.dictionary import Dictionary
 
@@ -96,10 +97,11 @@ def main(args):
 
     logging.debug('Now beginning VSM construction with Word2Vec')
 
-    model = Word2Vec(sentences=documents, vocab_path=args.vocab_path, workers=4,
+    model = Word2Vec(sentences=documents, vocab_path=args.vocab_path,
                      window=args.window_size,
                      min_count=args.minimum_token_count,
-                     size=args.vector_dimensions)
+                     size=args.vector_dimensions,
+                     workers=multiprocessing.cpu_count())
 
     model.save(args.out_path)
 

### 2014-05-13 vsm/corpora/wiki.py
corpora.wiki: set sentences out handle before super'ing

@@ -69,15 +69,14 @@ class WikiSentenceCorpus(WikiCorpus):
         dictionary_save_path = kwargs.pop('dictionary_save_path', None)
         sentences_save_path = kwargs.pop('sentences_save_path', None)
 
+        self.sentences_out = None
+        if sentences_save_path is not None:
+            self.sentences_out = open(sentences_save_path, 'w')
+
         super(WikiSentenceCorpus, self).__init__(*args, **kwargs)
 
         if dictionary_save_path is not None:
             self.dictionary.save(dictionary_save_path)
-
-        if sentences_save_path is not None:
-            self.sentences_out = open(sentences_save_path, 'w')
-        else:
-            self.sentences_out = None
 
     def open_corpus_file(self):
         _, extension = os.path.splitext(self.fname)

### 2014-05-13 tools/evaluate.py
evaluate: spawn a pool with size based on actual CPU count

@@ -6,6 +6,7 @@ import argparse
 import codecs
 from datetime import datetime
 import logging
+import multiprocessing
 from multiprocessing.pool import Pool
 import random
 import sys
@@ -76,7 +77,7 @@ def evaluate_model(model, data, do_train=True):
     global MODEL
     MODEL = model
 
-    pool = Pool(4)
+    pool = Pool(multiprocessing.cpu_count())
     scores = [x for x in pool.imap_unordered(score, test_pairs)
               if x is not None]
     logging.debug("Scores: %r" % scores)

### 2014-05-14 tools/evaluate.py
Restructure `evaluate` tool

@@ -5,6 +5,7 @@ data.
 import argparse
 import codecs
 from datetime import datetime
+import json
 import logging
 import multiprocessing
 from multiprocessing.pool import Pool
@@ -50,35 +51,37 @@ def score((source_word, target_word), threshold=20):
     return score
 
 
-def evaluate_model(model, data, do_train=True):
-    """Evaluate the performance of a translation model. Returns the mean
-    and standard deviation of scores among data tuples used for testing
-    (where a score ranges between 0 (worst) and 1 (perfect
-    translation))."""
+def save_model(model, originating_arguments):
+    """Save a newly trained model along with information about the
+    arguments that generated it.."""
 
-    if do_train:
-        training_pairs, test_pairs = train_test_split(data)
-        model.train(training_pairs)
+    now = datetime.now()
+    model_name = ("saved_models/model-{}-{}-{:02d}{:02d}"
+                  .format(model.__class__.__name__, now.date().isoformat(),
+                          now.hour, now.minute))
 
-        # We just trained a model -- save it somewhere
-        now = datetime.now()
-        model_name = ("saved_models/model-{}-{}-{:02d}{:02d}"
-                      .format(model.__class__.__name__, now.date().isoformat(),
-                              now.hour, now.minute))
-
-        try:
-            model.save(model_name)
-        except NotImplementedError:
-            logging.info("Model does not support saving to files; "
-                         "skipping save")
+    try:
+        model.save(model_name)
+    except NotImplementedError:
+        logging.info("Model does not support saving to files; "
+                     "skipping save")
     else:
-        test_pairs = data
+        # Save argument information as well
+        with open('{}_arguments.json', 'w') as arguments_f:
+            json.dump(originating_arguments, arguments_f)
+
+
+def evaluate_model(model, test_data):
+    """Evaluate the performance of a trained translation model. Returns
+    the mean and standard deviation of scores among data tuples used for
+    testing (where a score ranges between 0 (worst) and 1 (perfect
+    translation))."""
 
     global MODEL
     MODEL = model
 
     pool = Pool(multiprocessing.cpu_count())
-    scores = [x for x in pool.imap_unordered(score, test_pairs)
+    scores = [x for x in pool.imap_unordered(score, test_data)
               if x is not None]
[…]

### 2014-05-14 test/test_evaluate.py
Start tests for evaluate tool

@@ -1,0 +1,23 @@
+from nose.tools import assert_equal
+
+from tools import evaluate
+
+
+class DummyNamespace(object):
+    """Fake an argparse Namespace result"""
+
+    def __init__(self, entries):
+        self.__dict__.update(entries)
+
+
+def test_evaluate_dummy_neural():
+    arguments = {
+        'model': 'neural',
+        'vsm_source': 'test/fixtures/single-word/vsm-a-window5-min0-dim2',
+        'vsm_target': 'test/fixtures/single-word/vsm-b-window5-min0-dim2',
+        'data': 'test/fixtures/single-word/seed-set.tsv',
+        'model_arguments': {'hidden_layer_size': 2},
+        'model_file': None,
+    }
+
+    evaluate.main(DummyNamespace(arguments))

### 2014-05-14 tools/evaluate.py
evaluate: add flag to test on training data

@@ -114,6 +114,8 @@ def parse_args():
                         type=parse_cmdline_kwarg,
                         help=('Float or integer keyword arguments to '
                               'pass to the model (of the form )'))
+    parser.add_argument('--test-on-train', default=False,
+                        help='Test on the same data used to train')
 
     arguments = parser.parse_args()
 
@@ -156,7 +158,11 @@ def main(arguments):
                           "saved files")
             sys.exit(1)
     else:
-        training_pairs, test_pairs = train_test_split(data)
+        if arguments.test_on_train:
+            training_pairs = test_pairs
+        else:
+            training_pairs, test_pairs = train_test_split(data)
+
         model.train(training_pairs)
 
     # Now perform evaluation

### 2014-05-14 model/neural.py
Neural model: fix dataset key name

@@ -44,7 +44,7 @@ class NeuralTranslationModel(TranslationModel):
 
         dataset = SupervisedDataSet(input_size, output_size)
         dataset.setField('input', np.mat(source_vecs))
-        dataset.setField('output', np.mat(target_vecs))
+        dataset.setField('target', np.mat(target_vecs))
 
         trainer = BackpropTrainer(self.network, dataset,
                                   learningrate=self.learning_rate,

### 2014-05-14 test/test_evaluate.py
Update test fixtures

@@ -13,11 +13,12 @@ class DummyNamespace(object):
 def test_evaluate_dummy_neural():
     arguments = {
         'model': 'neural',
-        'vsm_source': 'test/fixtures/single-word/vsm-a-window5-min0-dim2',
-        'vsm_target': 'test/fixtures/single-word/vsm-b-window5-min0-dim2',
-        'data': 'test/fixtures/single-word/seed-set.tsv',
-        'model_arguments': {'hidden_layer_size': 2},
+        'vsm_source': 'test/fixtures/three-words/vsm-a-window5-min0-dim5',
+        'vsm_target': 'test/fixtures/three-words/vsm-b-window5-min0-dim5',
+        'data': 'test/fixtures/three-words/seed-set.tsv',
+        'model_arguments': {'hidden_layer_size': 5},
         'model_file': None,
+        'test_on_train': True,
     }
 
     evaluate.main(DummyNamespace(arguments))

### 2014-05-14 model/neural.py
Neural model: add `verbose` parameter

@@ -25,7 +25,7 @@ class NeuralTranslationModel(TranslationModel):
 
     def __init__(self, source_vsm, target_vsm, bias=BIAS,
                  hidden_layer_size=HIDDEN_LAYER_SIZE,
-                 learning_rate=LEARNING_RATE):
+                 learning_rate=LEARNING_RATE, verbose=False):
         """TODO document"""
 
         super(NeuralTranslationModel, self).__init__(source_vsm, target_vsm)
@@ -35,6 +35,7 @@ class NeuralTranslationModel(TranslationModel):
         self.bias = bias
         self.hidden_layer_size = hidden_layer_size
         self.learning_rate = learning_rate
+        self.verbose = verbose
 
     def train_vecs(self, source_vecs, target_vecs):
         input_size = self.source_vsm.layer1_size
@@ -48,7 +49,7 @@ class NeuralTranslationModel(TranslationModel):
 
         trainer = BackpropTrainer(self.network, dataset,
                                   learningrate=self.learning_rate,
-                                  verbose=True)
+                                  verbose=self.verbose)
         trainer.trainUntilConvergence()
 
     def load(self, path):

### 2014-05-14 model/runner.py
Move some of the more generic evaluation routines out of the script file and into the main source code

@@ -1,0 +1,55 @@
+"""Handles the training and evaluation of new translation models.
+
+- Trains models given a seed set and input vector spaces.
+- Performs parallelized evaluation of models given a test set."""
+
+import logging
+import multiprocessing
+from multiprocessing.pool import Pool
+
+
+# Global reference to the trained model being evaluated, shared among
+# processes
+MODEL = None
+
+
+def score((source_word, target_word), threshold=20):
+    """Test the given model on the given `(source_word, target_word)`
+    pair and return a score describing its performance.
+
+    Allow the model to make `threshold` number of guesses before
+    concluding that we have a score of 0.
+
+    The score is bounded between 0 (worst) and 1 (perfect
+    translation)."""
+
+    try:
+        predictions = MODEL.translate(source_word, n=threshold)
+    except ValueError:
+        # Translation error
+        logging.exception("Translation error")
+        return None
+
+    try:
+        rank = predictions.index(target_word)
+    except ValueError:
+        score = 0
+    else:
+        score = 1.0 / (rank + 1)
+
+    return score
+
+
+def evaluate_model(model, test_data):
+    """Evaluate the performance of a trained translation model. Returns
+    the scores among data tuples provided for testing (where a score
+    ranges between 0 (worst) and 1 (perfect translation))."""
+
+    global MODEL
+    MODEL = model
+
+    pool = Pool(multiprocessing.cpu_count())
+    scores = [x for x in pool.imap_unordered(score, test_data)
+              if x is not None]
+
+    return scores

### 2014-05-14 tools/evaluate.py

@@ -16,6 +16,7 @@ from gensim.models import Word2Vec
 from numpy import mean, std
 
 import model
+from model.runner import evaluate_model, score
 
 
 def train_test_split(data, training_pct=80):
@@ -24,31 +25,6 @@ def train_test_split(data, training_pct=80):
     random.shuffle(data)
     split = int(len(data) * (training_pct / 100.0))
     return data[:split], data[split:]
-
-
-MODEL = None
-def score((source_word, target_word), threshold=20):
-    """Test the given model on the given `(source_word, target_word)`
-    pair and return a score describing its performance.
-
-    The score is bounded between 0 (worst) and 1 (perfect
-    translation)."""
-
-    try:
-        predictions = MODEL.translate(source_word, n=threshold)
-    except ValueError:
-        # Translation error
-        logging.exception("Translation error")
-        return None
-
-    try:
-        rank = predictions.index(target_word)
-    except ValueError:
-        score = 0
-    else:
-        score = 1.0 / (rank + 1)
-
-    return score
 
 
 def save_model(model, originating_arguments):
@@ -69,23 +45,6 @@ def save_model(model, originating_arguments):
         # Save argument information as well
         with open('{}_arguments.json', 'w') as arguments_f:
             json.dump(originating_arguments, arguments_f)
-
-
-def evaluate_model(model, test_data):
-    """Evaluate the performance of a trained translation model. Returns
-    the mean and standard deviation of scores among data tuples used for
-    testing (where a score ranges between 0 (worst) and 1 (perfect
-    translation))."""
-
-    global MODEL
-    MODEL = model
-
-    pool = Pool(multiprocessing.cpu_count())
-    scores = [x for x in pool.imap_unordered(score, test_data)
-              if x is not None]
-    logging.debug("Scores: %r" % scores)
-
-    return mean(scores), std(scores)
 
 
 def parse_cmdline_kwarg(kwarg):
@@ -168,7 +127,8 @@ def main(arguments):
     # Now perform evaluation
     #
     # TODO print nicely
-    print evaluate_model(model, data)
+    scores = evaluate_model(model, data)
+    print mean(scores), std(scores)
 
 
 if __name__ == '__main__':

### 2014-05-14 test/test_evaluate.py
test_evaluate -> test_neural

@@ -1,24 +1,0 @@
-from nose.tools import assert_equal
-
-from tools import evaluate
-
-
-class DummyNamespace(object):
-    """Fake an argparse Namespace result"""
-
-    def __init__(self, entries):
-        self.__dict__.update(entries)
-
-
-def test_evaluate_dummy_neural():
-    arguments = {
-        'model': 'neural',
-        'vsm_source': 'test/fixtures/three-words/vsm-a-window5-min0-dim5',
-        'vsm_target': 'test/fixtures/three-words/vsm-b-window5-min0-dim5',
-        'data': 'test/fixtures/three-words/seed-set.tsv',
-        'model_arguments': {'hidden_layer_size': 5},
-        'model_file': None,
-        'test_on_train': True,
-    }
-
-    evaluate.main(DummyNamespace(arguments))

### 2014-05-14 test/test_neural.py

@@ -1,0 +1,21 @@
+from gensim.models import Word2Vec
+from nose.tools import assert_equal
+
+from model import NeuralTranslationModel
+from model.runner import evaluate_model
+
+
+def load_vsm_fixture(name):
+    return Word2Vec.load('test/fixtures/{}'.format(name))
+
+
+def test_3words():
+    data = [('here', 'there'), ('now', 'then')]
+
+    model = NeuralTranslationModel(
+        load_vsm_fixture('three-words/vsm-a-window5-min0-dim5'),
+        load_vsm_fixture('three-words/vsm-b-window5-min0-dim5'),
+        hidden_layer_size=5, learning_rate=0.00001)
+    model.train(data)
+
+    print evaluate_model(model, data)

### 2014-05-15 test/test_neural.py
test_neural updates

@@ -15,7 +15,8 @@ def test_3words():
     model = NeuralTranslationModel(
         load_vsm_fixture('three-words/vsm-a-window5-min0-dim5'),
         load_vsm_fixture('three-words/vsm-b-window5-min0-dim5'),
-        hidden_layer_size=5, learning_rate=0.00001)
+        hidden_layer_size=5, learning_rate=0.01)
     model.train(data)
 
+    # TODO how to assert good enough performance here? Maybe train an ensemble?
     print evaluate_model(model, data)

### 2014-05-15 test/test_linear.py
add test_linear

@@ -1,0 +1,21 @@
+from gensim.models import Word2Vec
+from nose.tools import assert_equal
+
+from model import LinearTranslationModel
+from model.runner import evaluate_model
+
+
+def load_vsm_fixture(name):
+    return Word2Vec.load('test/fixtures/{}'.format(name))
+
+
+def test_3words():
+    data = [('here', 'there'), ('now', 'then')]
+
+    model = LinearTranslationModel(
+        load_vsm_fixture('three-words/vsm-a-window5-min0-dim5'),
+        load_vsm_fixture('three-words/vsm-b-window5-min0-dim5'))
+    model.train(data)
+
+    for score in evaluate_model(model, data):
+        assert_equal(1.0, score, "Should see 100% performance on training data")

### 2014-05-15 model/core.py
model core: try fetching from VSM with some alternate encodings (ugh)

@@ -3,6 +3,28 @@ import logging
 from scipy.spatial import distance
 
 
+def get_word_vector(vsm, word, alternate_encodings=None):
+    """Retrieve the word vector for a given Unicode-encoded word from
+    a VSM.
+
+    Attempts to abstract away some encoding ugliness."""
+
+    if alternate_encodings is None:
+        alternate_encodings = 'latin-1'
+
+    try:
+        return vsm[word]
+    except KeyError:
+        for encoding in alternate_encodings:
+            try:
+                return vsm[word.encode(encoding)]
+            except UnicodeEncodeError:
+                pass
+
+        # Still here?
+        return None
+
+
 class TranslationModel(object):
     """A model which uses paired vector-space language models to translate
     words between two languages.
@@ -20,15 +42,13 @@ class TranslationModel(object):
 
         source_vecs, target_vecs = [], []
         for source_word, target_word in seeds:
-            try:
-                source = self.source_vsm[source_word]
-            except KeyError:
+            source = get_word_vector(self.source_vsm, source_word)
+            if source is None:
                 logging.warn(u'Source VSM missing word {}'.format(source_word))
                 continue
 
-            try:
-                target = self.target_vsm[target_word]
-            except KeyError:
+            target = get_word_vector(self.target_vsm, target_word)
+            if target is None:
                 logging.warn(u'Target VSM missing word {}'.format(target_word))
                 continue
 

### 2014-05-15 model/core.py
Fix typo

@@ -10,7 +10,7 @@ def get_word_vector(vsm, word, alternate_encodings=None):
     Attempts to abstract away some encoding ugliness."""
 
     if alternate_encodings is None:
-        alternate_encodings = 'latin-1'
+        alternate_encodings = ['latin-1']
 
     try:
         return vsm[word]

### 2014-05-15 model/core.py
VSM word fetch: ignore KeyError, too

@@ -18,8 +18,8 @@ def get_word_vector(vsm, word, alternate_encodings=None):
         for encoding in alternate_encodings:
             try:
                 return vsm[word.encode(encoding)]
-            except UnicodeEncodeError:
-                pass
+            except UnicodeEncodeError: pass
+            except KeyError: pass
 
         # Still here?
         return None

### 2014-05-15 model/core.py
VSM lookup: try utf-8 encoding too..

@@ -10,7 +10,7 @@ def get_word_vector(vsm, word, alternate_encodings=None):
     Attempts to abstract away some encoding ugliness."""
 
     if alternate_encodings is None:
-        alternate_encodings = ['latin-1']
+        alternate_encodings = ['utf-8', 'latin-1']
 
     try:
         return vsm[word]

### 2014-05-15 model/neural.py
First stab at pylearn2 implementation

@@ -1,11 +1,11 @@
 import logging
+import pickle
 
 import numpy as np
-from pybrain.datasets import SupervisedDataSet
-from pybrain.supervised.trainers import BackpropTrainer
-from pybrain.tools.shortcuts import buildNetwork
-from pybrain.tools.xml.networkreader import NetworkReader
-from pybrain.tools.xml.networkwriter import NetworkWriter
+from pylearn2.models import mlp
+from pylearn2.training_algorithms import sgd
+from pylearn2.termination_criteria import MonitorBased
+from pylearn2.datasets.dense_design_matrix import DenseDesignMatrix
 
 from model.core import TranslationModel
 
@@ -23,9 +23,12 @@ class NeuralTranslationModel(TranslationModel):
 
     LEARNING_RATE = 0.01
 
+    BATCH_SIZE = 10
+
     def __init__(self, source_vsm, target_vsm, bias=BIAS,
                  hidden_layer_size=HIDDEN_LAYER_SIZE,
-                 learning_rate=LEARNING_RATE, verbose=False):
+                 learning_rate=LEARNING_RATE, batch_size=BATCH_SIZE,
+                 verbose=False):
         """TODO document"""
 
         super(NeuralTranslationModel, self).__init__(source_vsm, target_vsm)
@@ -35,29 +38,53 @@ class NeuralTranslationModel(TranslationModel):
         self.bias = bias
         self.hidden_layer_size = hidden_layer_size
         self.learning_rate = learning_rate
+        self.batch_size = batch_size
         self.verbose = verbose
 
+    def foo(self, t):
+        print t.model.monitor.channels
+
     def train_vecs(self, source_vecs, target_vecs):
+        # Build dataset
+        X = np.mat(source_vecs)
+        Y = np.mat(target_vecs)
+        dataset = DenseDesignMatrix(X=X, y=Y)
+
+        # Determine visible layer dimensions
         input_size = self.source_vsm.layer1_size
         output_size = self.target_vsm.layer1_size
-        self.network = buildNetwork(input_size, self.hidden_layer_size,
-                                    output_size, bias=self.bias, fast=True)
 
-        dataset = SupervisedDataSet(input_size, output_size)
-        dataset.setField('input', np.mat(source_vecs))
-        dataset.setField('target', np.mat(target_vecs))
+        # Hidden layer with sigmoid activation function
+        hidden_layer = mlp.Sigmoid(layer_name='hidden', irange=.1, init_bias=1.,
[…]

### 2014-05-15 model/neural.py
Forgot argument

@@ -75,7 +75,7 @@ class NeuralTranslationModel(TranslationModel):
         self.network = mlp.MLP(layers, nvis=input_size)
         trainer.setup(self.network, dataset)
 
-        trainer.train()
+        trainer.train(dataset=dataset)
 
     def load(self, path):
         with open(path, 'r') as model_f:

### 2014-05-15 model/neural.py
Various pylearn2 fixes

@@ -1,10 +1,12 @@
 import logging
 import pickle
+from pprint import pprint
 
 import numpy as np
+from pylearn2.costs import mlp as mlp_costs
 from pylearn2.models import mlp
 from pylearn2.training_algorithms import sgd
-from pylearn2.termination_criteria import MonitorBased
+from pylearn2.termination_criteria import EpochCounter, MonitorBased
 from pylearn2.datasets.dense_design_matrix import DenseDesignMatrix
 
 from model.core import TranslationModel
@@ -41,9 +43,6 @@ class NeuralTranslationModel(TranslationModel):
         self.batch_size = batch_size
         self.verbose = verbose
 
-    def foo(self, t):
-        print t.model.monitor.channels
-
     def train_vecs(self, source_vecs, target_vecs):
         # Build dataset
         X = np.mat(source_vecs)
@@ -69,13 +68,20 @@ class NeuralTranslationModel(TranslationModel):
         trainer = sgd.SGD(learning_rate=self.learning_rate,
                           batch_size=self.batch_size,
                           termination_criterion=MonitorBased(),
-                          update_callbacks=self.foo)
+                          monitoring_dataset=dataset)
 
         # Now construct neural network
         self.network = mlp.MLP(layers, nvis=input_size)
         trainer.setup(self.network, dataset)
 
-        trainer.train(dataset=dataset)
+        while True:
+            trainer.train(dataset=dataset)
+
+            self.network.monitor.report_epoch()
+            self.network.monitor()
+
+            if not trainer.continue_learning(self.network):
+                break
 
     def load(self, path):
         with open(path, 'r') as model_f:

### 2014-05-21 paper/presentation/presentation.tex
add presentation (in progress)

\documentclass[color=usenames,dvipsnames]{beamer}

\usepackage{tikz}
\usepackage[utf8]{inputenc}

% smart quotes
\usepackage[autostyle, english=american]{csquotes}
\MakeOuterQuote{"}

% \usepackage{beamerthemesplit} // Activate for custom appearance

\title{Deep Neural Models\\for Bilingual Lexicon Extraction}
\author{Jon Gauthier and Arthur Tsang}
\date{28 May 2014}

\AtBeginSection[]
{
	\begin{frame}
		\tableofcontents[currentsection]
	\end{frame}
}

\begin{document}

\frame{\titlepage}

\section[Outline]{}
\frame{\tableofcontents}

\section{Introduction}
\subsection{Neural networks and translation}
\frame
{
	\frametitle{How about this one?}
	\scalebox{2}{$f(\text{English}) = \text{Spanish}$}
}

\frame{
\resizebox{\linewidth}{!}{
\begin{tikzpicture}[font=\LARGE]
  \onslide<1->{
  \draw[rotate=15] (0, 0) rectangle (5, 8.09);
  \node (dog) [rotate=15] at (2, 6) {dog};
  \node (book) [rotate=15] at (1, 4) {book};
  \node (school) [rotate=15] at (2.4, 2.5) {school};

  \draw[rotate=-15] (6, 3) rectangle (11, 11.09);
  \node (perro) [rotate=-15] at (9.5, 6.5) {perro};
  \node (libro) [rotate=-15] at (10.5, 4.5) {libro};
  \node (escuela) [rotate=-15] at (9.25, 2.9) {escuela};
  }

  \onslide<2->{
  \draw[-latex] (dog) to[out=15, in=170, looseness=1]  (perro);
  \draw[-latex] (book) to[out=15, in=170, looseness=1]  (libro);
  \draw[-latex] (school) to[out=15, in=175, looseness=1]  (escuela);
  }
\end{tikzpicture}%
}
}

\section{The task: Bilingual lexicon extraction}

\frame{
\frametitle{Bilingual lexicon extraction}

Given multilingual corpora (composed of a \emph{source language} and \emph{target language}),\\~\\
\pause
learn \textbf{translation mappings} of the form
\pause

\[(\text{\emph{source language word}}, \text{\emph{target language word}})\]
}

\subsection{Standard lexical learning}

\frame{
\frametitle{\insertsubsectionhead}
\begin{figure}
\centering
\scalebox{2}{
\begin{tikzpicture}
  \node (the) at (0, 2) {\strut the};
  \node (cat) at (1, 2) {\strut cat};
  \node (is) at (2, 2) {\strut is};
  \node (here) at (3, 2) {\strut here};

  \node (aqui) at (0, 0) {\strut aquí};
  \node (esta) at (1, 0) {\strut está};
  \node (el) at (2, 0) {\strut el};
  \node (gato) at (3, 0) {\strut gato};

  \onslide<2->{
  \draw (the) -- (el);
  \draw (cat) -- (gato);
  \draw (is) -- (esta);
  \draw (here) -- (aqui);
  }
\end{tikzpicture}
}
\end{figure}
}

\frame{
\frametitle{But we need parallel corpora for this..}
\begin{columns}
	\begin{column}{0.5\textwidth}
		{\color{Green} Soy siervo del Fuego Secreto, administrador de la llama de Anor.}
		{\color{Peach} ¡Tu fuego oscuro es en vano, llama de Udûn!}
		{\color{Sepia} Regresa a las sombras.}
		{\color{Bittersweet} ¡No puedes pasar!}
		{\color{CadetBlue} ¡Corred, insensatos!}
	\end{column}
	\begin{column}{0.5\textwidth}
		{\color{Green} I am a servant of the secret fire, wielder of the flame of Anor.}
		{\color{Peach} The dark fire will not avail you, flame of Udûn!}
		{\color{Sepia} Go back to the Shadow.}
		{\color{Bittersweet} You shall not pass!}
		{\color{CadetBlue} Fly, you fools!}
	\end{column}
\end{columns}
}

\subsection{Minimally-supervised approaches}

\frame{
\frametitle{Non-parallel (comparable) corpora}
\small{
\begin{columns}
	\begin{column}{0.5\textwidth}
		Las \textcolor{Bittersweet}{letras}, como elementos de los \textcolor{CadetBlue}{alfabetos}, tienen un orden prescrito. Esto se conoce como "orden \textcolor{WildStrawberry}{alfabético}," aunque la clasificación \textcolor{WildStrawberry}{alfabética} es la ciencia dedicada a la tarea compleja de ordenar y clasificar las \textcolor{Bittersweet}{letras} en los diferentes \textcolor{JungleGreen}{idiomas}.
	\end{column}
	\begin{column}{0.5\textwidth}
		A \textcolor{Bittersweet}{letter} is a grapheme (written character) in an \textcolor{WildStrawberry}{alphabetic} system of writing, such as the \textcolor{CadetBlue}{alphabet} of the Greek \textcolor{JungleGreen}{language} and its descendants.
	\end{column}
\end{columns}
}
}

\end{document}

### 2014-05-23 model/runner.py
fix Unicode logging error

@@ -37,7 +37,7 @@ def log_translations(source_word, target_word, translations):
     targets = [u'**{}**'.format(target) if target == source_word else target
                for target in translations]
     logging.info(u'Translations of {}: {}'
-                 .format(source_word, ' '.join(targets)))
+                 .format(source_word, u' '.join(targets)))
 
 
 def score(source_word, expected_target_word, translations):

### 2014-05-23 tools/evaluate.py
fix command line parsing

@@ -73,7 +73,7 @@ def parse_args():
                         type=parse_cmdline_kwarg,
                         help=('Float or integer keyword arguments to '
                               'pass to the model (of the form )'))
-    parser.add_argument('--test-on-train', default=False,
+    parser.add_argument('--test-on-train', default=False, action='store_true',
                         help='Test on the same data used to train')
 
     arguments = parser.parse_args()

### 2014-06-01 model/__init__.py
Add RandomTranslationModel

@@ -1,3 +1,4 @@
 from core import TranslationModel
 from linear import LinearTranslationModel
 from neural import NeuralTranslationModel
+from random import RandomTranslationModel

### 2014-06-01 model/random.py

@@ -1,0 +1,26 @@
+from numpy.random import random
+
+from model import TranslationModel
+
+
+class RandomTranslationModel(TranslationModel):
+    # Lower and upper bounds for randomly generated values
+    LOWER_BOUND = -1.8777409
+    UPPER_BOUND = 1.8641127
+
+    def __init__(self, source_vsm, target_vsm, lower_bound=LOWER_BOUND,
+                 upper_bound=UPPER_BOUND):
+        super(RandomTranslationModel, self).__init__(source_vsm, target_vsm)
+
+        self.dim = self.target_vsm.layer1_size
+        self.lower_bound = lower_bound
+        self.upper_bound = upper_bound
+
+    def train_vecs(self, source_vecs, target_vecs):
+        # No training necessary!
+        pass
+
+    def translate_vec(self, source_vec):
+        return ((self.upper_bound - self.lower_bound)
+                * random(self.dim)
+                + self.lower_bound)

### 2014-06-01 tools/evaluate.py

@@ -89,6 +89,7 @@ def parse_args():
 MODEL_MAPPING = {
     'linear': model.LinearTranslationModel,
     'neural': model.NeuralTranslationModel,
+    'random': model.RandomTranslationModel,
 }
 
 def main(arguments):

### 2014-06-02 model/__init__.py
Add IdentityTranslationModel, PercentileFrequencyTranslationModel

@@ -1,4 +1,7 @@
 from core import TranslationModel
+
+from identity import IdentityTranslationModel
 from linear import LinearTranslationModel
 from neural import NeuralTranslationModel
+from percentile_frequency import PercentileFrequencyTranslationModel
 from random import RandomTranslationModel

### 2014-06-02 model/identity.py

@@ -1,0 +1,17 @@
+from model import TranslationModel
+
+
+class IdentityTranslationModel(TranslationModel):
+    """Treats the translation of any input vector as exactly that input
+    vector. (Assumes the dimensionality of the input and output spaces
+    match.)"""
+
+    def __init__(self, source_vsm, target_vsm):
+        super(IdentityTranslationModel, self).__init__(source_vsm, target_vsm)
+
+    def train_vecs(self, source_vecs, target_vecs):
+        # No training necessary!
+        pass
+
+    def translate_vec(self, source_vec):
+        return source_vec

### 2014-06-02 model/percentile_frequency.py

@@ -1,0 +1,43 @@
+import operator
+
+from model import TranslationModel
+
+
+class PercentileFrequencyTranslationModel(TranslationModel):
+    """Attempts to match input words with words in the target language
+    which have similar percentile corpus frequencies."""
+
+    def __init__(self, source_vsm, target_vsm):
+        super(PercentileFrequencyTranslationModel, self).__init__(source_vsm,
+                                                                  target_vsm)
+
+        self.source_distr = self.build_percentile_distribution(source_vsm)
+        self.target_distr = self.build_percentile_distribution(target_vsm)
+
+    def build_percentile_distribution(self, vsm):
+        """Build a word/corpus-frequency-percentile distribution given
+        a VSM."""
+
+        # Map word to corpus frequency, sort by increasing frequency
+        freqs = sorted(((word, vocab.count)
+                        for word, vocab in vsm.vocab.iteritems()),
+                       key=operator.itemgetter(1))
+
+        vocab_size = float(len(vsm.vocab))
+        percentiles = {w: i / vocab_size for i, (w, _) in enumerate(freqs)}
+
+        return percentiles
+
+    def train_vecs(self, source_vecs, target_vecs):
+        pass
+
+    def translate(self, word, n=5):
+        try:
+            source_percentile = self.source_distr[word]
+        except KeyError:
+            raise ValueError(u"Word '{}' not found in source VSM".format(word))
+
+        target_words = sorted(self.target_distr,
+                              key=lambda w: abs(self.target_distr[w]
+                                                - source_percentile))
+        return target_words[:n]

### 2014-06-02 model/runner.py

@@ -15,7 +15,7 @@ MODEL = None
 logger = logging.getLogger(__name__)
 
 
-def get_translations(source_word, threshold=100):
+def get_translations(source_word, threshold=10000):
     """Fetch the top `threshold` translations for the given
     source-language word.
 

### 2014-06-02 tools/evaluate.py

@@ -87,8 +87,10 @@ def parse_args():
 
 
 MODEL_MAPPING = {
+    'identity': model.IdentityTranslationModel,
     'linear': model.LinearTranslationModel,
     'neural': model.NeuralTranslationModel,
+    'percentile_frequency': model.PercentileFrequencyTranslationModel,
     'random': model.RandomTranslationModel,
 }
 

### 2014-06-03 paper/main.tex
Add paper

@@
 \usepackage{commands}
+\usepackage{cleveref}
+\usepackage{placeins} % FloatBarrier
+
+\usepackage{tikz}
+\usepackage{pgfplots}
+
+% DEV
+\usepackage{booktabs}
 
@@
 
-\title{Deep Neural Models for Bilingual Lexicon Extraction}
+\title{Deep Neural Networks for Bilingual Lexicon Extraction}
 
-\author{Jon Gauthier \\
-  Symbolic Systems Program \\
-  Stanford University \\
-  Stanford, CA 94305 \\
-  {\tt jgauthie@stanford.edu} \\\And
-  Arthur Tsang \\
-  Department of Computer Science \\
-  Stanford University \\
-  Stanford, CA 94305 \\
-  {\tt atsang2@stanford.edu} \\}
+%\author{Jon Gauthier\affilA\,and\,Arthur Tsang\affilB \\
+%  \affilA Symbolic Systems, \affilB Computer Science \\
+%Stanford University, Stanford, CA 94035 \\
+%{\tt \{jgauthie,atsang2\}@stanford.edu}}
 
@@
 \begin{abstract}
-  We present a new method for producing bilingual lexicons from very
-  non-parallel corpora. In contrast with previous approaches, our method
-  separates the steps of \begin{inparaenum}[(1)]
+  We explore a method for producing bilingual lexicons from non-parallel
+  corpora. In contrast with most previous approaches, our method separates the steps
+  of \begin{inparaenum}[(1)]
     \item constructing distributed word representations from monolingual
@@
   \end{inparaenum}
-
-  % TODO
-  We demonstrate that a deep neural model can effectively learn a translational
-  mapping between two monolingual vector spaces for the purposes of bilingual
-  lexicon extraction. We evaluate the performance of this model relative to X
-  and Y.
+  This gives us precise control over the independent development of the word representations and the mapping between them.
+  %By performing these two steps separately, we gain great flexibility in determining how our word representations are constructed.
+% are able to construct distributed word representations with greater flexibility.
+% TODO more specific and contentful
+  We demonstrate that a deep neural network can effectively learn a translational
+  mapping between these representations for the purposes of bilingual lexicon extraction more accurately than a linear model.
+  %explain baseline
+  %We suggest that a deep learning model beats a linear baseline by allowing more flexibility. % make this more concrete
 \end{abstract}
@@
 \section{Introduction}
+\label{sec:introduction}
 
-In the task of bilingual lexicon extraction (BLE), we accept as input some pair
-of corpora in different languages---perhaps parallel or perhaps not---and learn
-associations of translational equivalence between one language (a "source
-language") and the other (a "target language").
+In bilingual lexicon extraction (BLE), we accept as input some pair of corpora
+in different languages---perhaps parallel or perhaps not---and use them to learn associations of translational equivalence between a \textit{source language} and a \textit{target language}.
 
@@
 under-resourced language pairs which do not already benefit from a surplus of
-lexicon materials. These same language pairs, lacking resources as simple as
-translation dictionaries, also lack sufficient aligned text for training
-BLE models. For this reason, most recent work on this task focuses on
+lexicon materials. These same language pairs, lacking resources as simple as quality translation dictionaries, also lack sufficient amounts of manually aligned
+text to train BLE models. % briefly mention that each individual language may have a significant corpus though?
+For this reason, most recent work on this task focuses on
 minimally supervised approaches which do not require parallel corpora
@@
 of word co-occurrences in different languages." In line with this hypothesis,
-recent work has determined the likelihood of two words as translations of one
-another by comparing the translations of the contexts in which they appear. The
-standard approach is to manually\unclear construct a "bilingual vector space" which
-contains distributed word representations of words in both the source and target
-language \cite{fung1998,peirsman2010,vulic2013}.
+recent work has related the likelihood of two words being translations of one
+another to the similarities between the contexts in which the words appear in
+their corresponding languages. The standard approach under this theory is to
+manually construct a \textit{bilingual vector space} which contains distributed
+representations of words in both the source and target language
+\cite{fung1998,vulic2013}. The axes of this space are built from
+bilingual seed pairs. A given word representation has a large value along an axis
+if it co-occurs often with the seed word corresponding to that axis. Because
+axes are defined with bilingual pairs, word representations for both source and
+target language words can lie in the same space. With a bilingual vector space
+constructed, translations for a given source-language word can be retrieved by
+simply finding the nearest target-language word representation neighboring the
+word's own vector representation in the bilingual vector space.
 
-\section*{Acknowledgments}
+The present paper is motivated by the hypothesis that this standard bilingual
+vector space approach to BLE is inherently limited by the design of the space
+itself.
+% SHORT We suggest that the construction of the space along axes composed of a hand-picked set of bilingual seeds has the potential to limit the expressiveness of the word representations contained therein.
+Consider, for example, a vector
+space constructed in this fashion using just two seeds: \textit{hot} and
+\textit{cold}. Such a space may be effective at representing the difference
+between words like \textit{fire} and \textit{ice}. However, with these two axes
+it would fail to capture many other distinctions, such as the difference between
+the words \textit{up} and \textit{down}. The quality of the vector space is
+thus heavily dependent on the quality of its seeds.
 
-The acknowledgments should go immediately before the references.  Do
-not number the acknowledgments section. Do not include this section
-when submitting your paper for review.
+\newcite{mikolov2013exploiting} suggest an alternate method for bilingual lexicon
+extraction, centered around two monolingual vector spaces $E$ (corresponding to the
+source language) and $F$ (corresponding to the target language). These vector spaces
+are constructed using a standard neural language model \cite{mikolov2013}.
+We see this as a promising alternative to the aforementioned bilingual vector space approach, as it allows for the construction of robust distributed representations whose quality is not dependent on manually selected seeds.
+The authors find
+that a linear model trained via stochastic gradient descent to map between these two
+vector spaces can effectively map source-language word representations to the
+target-language word vectors corresponding to their translations.
+
+We extend this work in attempting to model a \emph{translation function} $t : E \to F$.
+% SHORT
+% We capitalize on the fact that these vector spaces may be independently constructed, experimenting with different configurations for each space.
+In this paper we approximate the transformation $t$ between the two spaces using a deep neural
+network, trained to accept as input word vectors from the source language space
+$E$ and output word vectors in the target language space $F$.
+
+% \Cref{sec:data} describes the corpora used to test this translation process.
+% \Cref{sec:model} details the actual process of translation. In
+% \Cref{sec:evaluation,sec:conclusion} we analyze the performance of our model and
+% examine how it supports our motivating hypothesis.
+
+\section{Data}
+\label{sec:data}
+
+Our translation algorithm makes use of two monolingual corpora: one consisting of text
+in the source language and the other containing text in the target language. We presume
+the corpora have the following properties:
+\subparagraph{Large and broad-domain.} With this property given, we can safely assume
+that for any common
+word which appears in the source-language corpus, we will also find its
+translation somewhere in the target-language corpus.
+If large corpora are unavailable, we can make use of smaller narrow-domain corpora, such as religious texts.
+\subparagraph{Non-parallel.} This means that our algorithm does not depend on any sort
+of alignment, lexical or topical. This relative flexibility allows us to demonstrate
+that the method can be applied in situations where parallel corpora are not available.
+
+% TODO for long paper: describe preprocessing techniques (stolen from Gensim).
+In our experiments we use English and Spanish corpora sourced from Wikipedia.%
+\footnote{Available at \url{http://dumps.wikimedia.org/}.}
+We choose not to take advantage of the document alignments available in the Wikipedia corpus in order to show that our method is generalizable to other corpora.
+
+In addition to the two monolingual corpora, we train on a minimal
+%, manually-created
+set of word translation pairs.
+% TODO I don't like "such as a swadesh list" --- Jon
+The specifics of these training pairs are described in \Cref{subsec:translation-nn}.
+
+%In addition to using the corpora, we also rely on a small seed set of several hundred English % -Spanish word pairs to train the translation function.
+% Include the seed sets as an appendix?
+
+\section{Model}
+\label{sec:model}
+
+\begin{figure}[tb!]
+	\centering
+	\resizebox{\linewidth}{!}{
+    \begin{tikzpicture}
+        \node (english) at (0, 2) {\large \textbf{book}};
+
+        \draw (3.25, 2) ellipse[x radius=0.75, y radius=1.1];
+        \node (englishvec) at (3.25, 2) {$\begin{bmatrix}~\vdots~\\\vdots\end{bmatrix}$};
+
+        \draw (6.5, 2) ellipse[x radius=0.75, y radius=1.1];
+        \node (spanishvec) at (6.5, 2) {$\begin{bmatrix}~\vdots~\\\vdots\end{bmatrix}$};
+
+        \node (spanish) at (9.5, 2) {\large \textbf{libro}};
+
+        \draw [->] (english) -- (englishvec) node [pos=0.45, above] {\small \texttt{word2vec}};
+        \draw [->] (englishvec) -- (spanishvec) node [midway, above, align=center] {Neural\\network};
+        \draw [->] (spanishvec) -- (spanish) node [pos=0.55, above, align=center] {Nearest\\neighbor};
+    \end{tikzpicture}
+    }
+    \caption{The process of word translation, using a neural network to map between
+        two monolingual VSMs}
+    \label{fig:translation}
+\end{figure}
+
+\Cref{fig:translation} illustrates the process of translating a single word from a source
+language to a target language (here English and Spanish, respectively) in this model. We
+begin with an input source-language word and retrieve a corresponding pre-computed word
+representation in the source VSM. We use this source-language word vector as an input to
+a neural network, which outputs a vector in the target-language VSM. We perform a simple
+nearest-neighbor search in the target-language VSM to find the closest existing word
+vector, and take its corresponding target-language word to be the translation of the
+original source-language word.
+
+\subsection{Word representations}
+\label{subsec:word-representations}
+
+For each of the two corpora introduced in \Cref{sec:data}, we train a continuous vector space model on 10-word contexts, using an implementation in the \textit{Gensim} software package of \texttt{word2vec}, a neural language model \cite{rehurek2010,mikolov2013}.
+This neural language model attempts to learn word representations which minimize error in the task of predicting a word's context given the word itself.
+The resulting word vectors are thus theoretically optimal (i.e., minimum-cost) representations of the meanings underlying their corresponding words.
+This is in contrast with the word representations built by a standard BLE approach, which define themselves in terms of manually constructed axes, and thus do not necessarily account for a maximal amount of variance in meaning among the words represented.
+
+We experiment with different word vector dimensionalities, and detail our specific choices later in \Cref{subsec:test-data}.
+
+%A VSM is structured such that the relative position of a word encodes aspects of its meaning.
+%For example, similar words map into similar neighborhoods, and addition by the same vector can produce analogies (citation needed).\unnec
+% Ok, I suppose we don't need to give that much detail (especially for a 4 page paper).
+A brief mention, though, can help explain the intuition that some structure can be extrapolated by a neural network --- Arthur
+% Since independently-produced VSMs maintain similar patterns, this suggests we can build a translation function from a small number of seeds.
+
+\subsection{Translation with a neural network}
+\label{subsec:translation-nn}
+
+The computation central to the process of translation is the transformation of a
+source-language word representation into a target-language word representation via
+a deep neural network. We learn this transformation from a small set of training pairs.
+%broad-domain training pairs, each mapping a single source language word to a single
+%translationally equivalent
+%target language word.
+
+% TODO added this.. what do you think? --- Jon
+The standard bilingual vector space approach detailed in the introduction would
+base its word representations on these training pairs. In contrast with this method,
+our model uses these translation pairs only in its final learning step, while training
+the neural network which models the translation function. For this reason, the model
+can benefit from effective word representations developed independently from the
+translation system itself. Furthermore, it can provide direct feedback through a training cost function
+on the quality of the translation pairs.
+
+We investigate several different methods 
+of producing translation pairs.
+In line with previous
+approaches, we fill a set
+of translation pairs
+\texttt{Cog} with cognates shared between the two
+languages \cite{koehn2002}. We also build a small separate set
+\texttt{ConcNN}
+consisting of concrete nouns, and attempt to restrict the set to word pairs for which
+both the source- and target-language words have just one sense in their respective
+languages.
+
+For each training pair of words $(e, f)$, we provide as a training example the input and output $(w_e, w_f)$, where $w_e$ and $w_f$ are word representations
+drawn from their respective VSMs constructed as described in \Cref{subsec:word-representations}.
+
+At inference time, for a given English word $e$, we provide as input to the neural network
+the corresponding word representation $w_e$. The output of the network, $w_f$, is some
+vector in the target-language vector space. We search for the nearest known neighbor of
+$w_f$, and treat the word corresponding to this nearest neighbor as the translation $f$
+of the input word $e$.
+
+\subsection{Comparison: Translation with a linear transformation}
+\label{subsec:translation-linear}
+
+\newcite{mikolov2013exploiting} found that a linear model could learn a translation mapping
+between two monolingual vector spaces. We construct such a linear model for purposes of
+comparison with the neural network method detailed above. It utilizes
+the same basic idea as presented in \Cref{fig:translation}: it only differs in the type of
+transformation used to map between the two monolingual vector spaces. Whereas our approach uses
+a deep neural network, this linear model builds a linear transformation between the two spaces.
+Given source-language and target-language training data matrices $W_E, W_F$ (where the $i$th
+columns of the two matrices are the vector word representations corresponding to the $i$th
+training pair) we assert the existence of a translation matrix $A$, where $W_F = AW_E$.
+
+We compute the linear transformation by taking $A = W_F W_E^+$, where $W_E^+$ is the Moore--Penrose
+pseudoinverse of $W_E$.
+% SHORT
+% It is necessary to use the pseudoinverse of $W_E$, as the matrix is unlikely to be square and invertible.
+Then given any source-language word vector $w_e$, we calculate
+the transformed target-language word vector as $w_f = A w_e$.
+
+\section{Evaluation}
+\label{sec:evaluation}
+% TODO maybe for long paper: note that test sets mirror the style of the seed set. e.g.
+% model trained on concrete nouns is tested on unseen concrete nouns; model tested on
+% cognates is tested on unseen cognates
+We evaluate our translation model on an unseen set of English-Spanish word pairs.
+Since our model builds a ranked list of nearest-neighbor translation candidates, we use the mean reciprocal rank (MRR) of the correct translation within the list:
+\begin{equation}
+    \textit{MRR} = \frac{1}{n}\sum_{j=1}^n \frac{1}{r_j}
+\end{equation}
+where $r_j$ is the rank of the correct translation for pair $j$ out of $n$.
+%For example, if every correct translation is ranked second highest, $\textit{MRR} = \frac{1}{2}$. % leaving out for space considerations
+
+\subsection{Baseline}
+\label{subsec:baseline}
+
+We establish a lower bound for translation performance with an algorithm that associates
+source- and target-language words by their relative corpus frequencies. Given an input
+source-language word, we compute its corpus frequency percentile. We then predict
+translations in the target language by returning the words with the closest percentile
+frequencies in the target language corpus.
+
+\subsection{Test data}
+\label{subsec:test-data}
+
+We establish three separate test environments, each defined by a choice of training data,
+input corpus format and VSM design. The vector space dimensionalities given below
+were derived through trial and error.
+% SHORT We find that different VSM dimensionalities are optimal for different corpus inputs.
+Overall, we see that performance varies significantly based on these test environment properties.
+The environments are as follows:
+\begin{description}
+\item[\texttt{Cog}:] Minimally preprocessed corpora, \texttt{Cog} training data, 500-dimensional word vectors % TODO long paper: 5000/750 min word threshold
+% Perhaps note that we don't do any Levenshtein mumbo-jumbo? --- Arthur
+\item[\texttt{ConcNN}:] Minimally preprocessed corpora, \texttt{ConcNN} training data, 500-dimensional word vectors % TODO long paper: 10000/1500 min word threshold
+\item[\texttt{CogLem}:] Lemmatized Spanish corpus, \texttt{Cog} training data, 1000-dimensional word vectors % TODO long paper: 2500/375 min word threshold
+\end{description}
+
+\begin{table*}[htb!]
+    \centering
+    \small
+    \begin{tabular}[t]{l lll}
+    \toprule
+    Diagnosis & Input & Expected cognate output & Predicted ranked output (translations) \\
+    \midrule
+    Synonyms & \emph{acceptance} & \emph{aceptación} & \emph{aprobación (approval), \dots} \\
+    Near misses & \emph{vanilla} & \emph{vainilla} & \emph{chocolate, vainilla, \dots} \\
+    Right neighborhood, wrong spot & \emph{election} & \emph{elección} & \emph{candidato, candidatura (candidacy), \dots} \\
+    Alternate sense interpretations & \emph{division} & \emph{división} & \emph{unidad (unit), élite, comando (command), \dots} \\
+    \bottomrule
+    \end{tabular}
+    \caption{Examples of common error patterns observed in the algorithm's output for
+        the \bt{Cog} environment}
+    \label{tbl:errors}
+\end{table*}
+
+Our evaluation metric looks for a single possible translation of a test input word in
+ranked output. As we explain in \Cref{subsec:error-analysis}, this may needlessly penalize
+the algorithm.
+% SHORT , as it often suggests reasonable translations earlier in the ranked output which happen to not match our strict search requirements.
+For example, our algorithm is
+tested on a cognate pair $(\textit{concept}, \textit{concepto})$, and is penalized for
+suggesting the translation \textit{idea}, a suitable synonym of the desired target word.
+A fourth test run \bt{CogFix} considers the algorithm's output on the \texttt{Cog} dataset,
+and accepts such sensible alternate translations. % SHORT (rather than stipulating in this case that all source-language cognates translate to their corresponding target-language cognates).
+
+\subsection{Results}
+\label{subsec:results}
+
+\begin{table}[tb]
+    \centering
+%    \begin{subtable}%{0.44\linewidth}
+        \resizebox{\linewidth}{!}{
+        \begin{tabular}[t]{l cccc}
+            % CogFix: manual fixes (allow plurals, synonyms, etc.)
+            \toprule
+            Model & \bt{Cog} & \bt{ConcNN} & \bt{CogLem} & \bt{CogFix} \\
+            \midrule
+            NN (125) & 0.216 & 0.199 & 0.215 & 0.389 \\
+            NN (250) & \textbf{0.295} & 0.193 & 0.269 & \textbf{0.480} \\
+            NN (500) & 0.264 & \textbf{0.203} & 0.279 & 0.432 \\
+            NN (1000) & 0.263 & 0.196 & 0.265 & 0.451 \\[1.5ex]
+            Linear & 0.280 & 0.195 & \textbf{0.286} & 0.449 \\
+            Baseline & 0.006 & 0.002 & 0.002 & 0.006 \\
+            \bottomrule
+        \end{tabular}
+        }
+%    \end{subtable}
+    \caption{Performance on multiple test environments (hidden layer dimensionalities given in parentheses). MRR values (higher is better).}
+    \label{tbl:results}
+%    \hfill
+%    \begin{subfigure}{0.55\linewidth}
+%        \resizebox{\linewidth}{!}{
+%        \begin{tikzpicture}
+%            \begin{axis}[xlabel=Hidden layer dimensionality, ylabel=MRR, legend pos=outer north east,
+%              legend style={font=\small}]
+%                \addplot[red, mark=x] coordinates {(125, 0.2158) (250, 0.2945) (500, 0.2635) (1000, 0.2628)};
+%                \addlegendentry{Cog}
+%                \addplot[red, dashed, forget plot, domain=125:1000] {0.2803};
+%
+%                \addplot[gray, mark=square*] coordinates {(125, 0.1991) (250, 0.1933) (500, 0.2029) (1000, 0.1956)};
+%                \addlegendentry{ConcNN}
+%                \addplot[gray, dashed, forget plot, domain=125:1000] {0.1954};
+%
+%                \addplot[blue, mark=*] coordinates {(125, 0.2147) (250, 0.2694) (500, 0.2788) (1000, 0.2646)};
+%                \addlegendentry{CogLem}
+%                \addplot[blue, dashed, forget plot, domain=125:1000] {0.2861};
+%
+%                \addplot[brown, mark=triangle*] coordinates {(125, 0.3891) (250, 0.4795) (500, 0.4318) (1000, 0.4509)};
+%                \addlegendentry{CogFix}
+%                \addplot[brown, dashed, forget plot, domain=125:1000] {0.4494};
+%          \end{axis}
+%        \end{tikzpicture}
+%        }
+%        \subcaption{Mean reciprocal ranks for various training environments (higher is better). Baseline performance for each seed set is illustrated as a flat dashed line in matching color.}
+%    \end{subfigure}
+%    \caption{Evaluation results}
+%    \label{fig:results}
+\end{table}
+
+Mean reciprocal rank values for each test environment and a number of training
+conditions are in \Cref{tbl:results}.
+The linear transformation model based on \newcite{mikolov2013exploiting} was very competitive with the neural network model.
+This suggests that many---but not all---of the geometric relationships in the target language vector space can be modeled through a simple linear transform of the geometric structures in the source language space.
+
+The best absolute performance is achieved by a neural network with 250 hidden-layer
+neurons in the \bt{CogFix} training environment. Interestingly, the hidden layer of this model is of
+lower dimensionality than the test input and output word vectors (each 500 dimensions).
+% TODO repetitive "This suggests" --- Jon
+% Mention that the Linear model may be overfitting, if compressing first leads to better results? --- Arthur
+This suggests that the model learns some useful compressed intermediate representation of
+the input source-language word in the translation training process.
+% SHORTCAND
+Its MRR shows us that on average the algorithm suggests an acceptable translation in the second or third position of the ranked list of possible translations.
+
+\subsection{Error analysis}
+\label{subsec:error-analysis}
+
+While the translation algorithm fails to predict the exactly correct translation a
+significant amount of the time, an analysis of common prediction errors is encouraging.
+We identify several common error patterns in the
+algorithm's output in \Cref{tbl:errors}, and elaborate on the classes of errors below:
+\subparagraph{Synonyms.} With many of the "errors" of this form, the algorithm provided an acceptable alternate translation that we simply did
+not expect. These cases are treated as successes in the \bt{CogFix} training environment, as described
+in \Cref{subsec:results}.
+\subparagraph{Near misses.} In these cases the algorithm mapped into a precise neighborhood
+of related words, but picked the wrong element from the set. % SHORT
+% For example, the output word \textit{chocolate} is strongly related to the expected word \textit{vainilla}.
+% TODO long paper: future work might examine the actual distance values and "give up" if the top few words are all very close
+\subparagraph{Right neighborhood, wrong spot.} We occasionally mapped into a correct larger
+area of the target language vector space, but still made a significant semantic error.
+% SHORT
+% With the input word \textit{election}, for example, the algorithm predicts as translations target-language words related to political campaigns.
+\subparagraph{Alternate senses.} In several cases on the \texttt{Cog} training
+set and its derivatives, the algorithm learned to predict translations of the input words which
+corresponded to valid alternate interpretations of the source language word.
+% SHORT
+% For example, it interprets the English \textit{division} in the sense of a military unit (rather than in the mathematical sense as we expected).
+Note that these cases were not treated as
+successes in the \bt{CogFix} training environment. We hope to eliminate this class of errors
+in future work by using multi-prototype VSMs (see \Cref{sec:conclusion}).
+
+% TODO long paper: Show T-SNE visualization of the neighborhoods into which we map erroneously --- show that we still map into the correct neighborhood, just the wrong spot, a lot of the time
+
+\section{Conclusion}
+\label{sec:conclusion}
+
+In this paper, we presented an algorithm which learns word translations from unlabeled,
+non-parallel corpora by building a mapping between two independently constructed
+monolingual vector spaces. This algorithm can be used to perform bilingual lexicon
+extraction in scenarios where the source and target language in question do not benefit
+from widely available parallel corpora. We demonstrated encouraging translation
+performance with large, broad-domain corpora as text sources.
+
+For simplicity, our current VSM design associates a single word vector with each word
+in the source or target language corpus. However, training error could be greatly
+reduced by allowing polysemous words to be associated with multiple word vectors
+\cite{reisinger2010,huang2012}. We believe such an extension will remove much of the remaining
+ambiguity in the input vector spaces.
+
+% However, as we note that many of our errors are caused by alternate sense interpretations, our approach could be further improved through disambiguation and multiple vector representations per word.
 

### 2014-06-07 paper/main.tex
Clean up paper LaTeX, begin extending for long paper version

@@
 
-%\author{Jon Gauthier\affilA\,and\,Arthur Tsang\affilB \\
-%  \affilA Symbolic Systems, \affilB Computer Science \\
-%Stanford University, Stanford, CA 94035 \\
-%{\tt \{jgauthie,atsang2\}@stanford.edu}}
+\author{Jon Gauthier,\affilA{} Arthur Tsang,\affilB{} Christopher Potts\affilC{} \\
+  \affilA{} Symbolic Systems Program, \affilB{}Department of Computer Science, \affilC{}Department of Linguistics \\
+Stanford University, Stanford, CA 94035 \\
+{\tt \{jgauthie,atsang2,cgpotts\}@stanford.edu}}
 
@@
   We explore a method for producing bilingual lexicons from non-parallel
-  corpora. In contrast with most previous approaches, our method separates the steps
-  of \begin{inparaenum}[(1)]
-    \item constructing distributed word representations from monolingual
-      corpora, and
-    \item building a mapping between these representations.
+  corpora. In contrast with most previous approaches, our method separates the
+  steps of \begin{inparaenum}[(1)]
+  \item constructing distributed word representations from monolingual corpora,
+    and
+  \item building a translational mapping.
   \end{inparaenum}
-  This gives us precise control over the independent development of the word representations and the mapping between them.
-  %By performing these two steps separately, we gain great flexibility in determining how our word representations are constructed.
-% are able to construct distributed word representations with greater flexibility.
-% TODO more specific and contentful
-  We demonstrate that a deep neural network can effectively learn a translational
-  mapping between these representations for the purposes of bilingual lexicon extraction more accurately than a linear model.
-  %explain baseline
-  %We suggest that a deep learning model beats a linear baseline by allowing more flexibility. % make this more concrete
+
+  This gives us precise control over the independent development of the word
+  representations and the mapping between them.
+
+  We demonstrate that a deep neural network can effectively learn a
+  translational mapping between these representations for the purposes of
+  bilingual lexicon extraction more accurately than a closely related linear
+  model.
 \end{abstract}
@@
 In bilingual lexicon extraction (BLE), we accept as input some pair of corpora
-in different languages---perhaps parallel or perhaps not---and use them to learn associations of translational equivalence between a \textit{source language} and a \textit{target language}.
+in different languages---perhaps parallel or perhaps not---and use them to learn
+associations of translational equivalence between a \textit{source language} and
+a \textit{target language}.
 
-Bilingual lexicon extraction is obviously of most utility when applied to
-under-resourced language pairs which do not already benefit from a surplus of
-lexicon materials. These same language pairs, lacking resources as simple as quality translation dictionaries, also lack sufficient amounts of manually aligned
-text to train BLE models. % briefly mention that each individual language may have a significant corpus though?
-For this reason, most recent work on this task focuses on
-minimally supervised approaches which do not require parallel corpora
+Bilingual lexicon extraction is of most utility when applied to under-resourced
+language pairs which do not already benefit from a surplus of lexicon materials.
+These same language pairs, lacking resources as simple as quality translation
+dictionaries, also lack sufficient amounts of manually aligned text to train BLE
+models. For this reason, most recent work on this task focuses on minimally
+supervised approaches which do not require parallel corpora
 \cite{rapp1995,peirsman2010}.
@@
 vector space approach to BLE is inherently limited by the design of the space
-itself.
-% SHORT We suggest that the construction of the space along axes composed of a hand-picked set of bilingual seeds has the potential to limit the expressiveness of the word representations contained therein.
-Consider, for example, a vector
+itself. We suggest that the construction of the space along axes composed of a
+hand-picked set of bilingual seeds has the potential to limit the expressiveness
+of the word representations contained therein. Consider, for example, a vector
 space constructed in this fashion using just two seeds: \textit{hot} and
@@
 it would fail to capture many other distinctions, such as the difference between
-the words \textit{up} and \textit{down}. The quality of the vector space is
-thus heavily dependent on the quality of its seeds.
+the words \textit{up} and \textit{down}. The quality of the vector space is thus
+heavily dependent on the quality of its seeds.
 
-\newcite{mikolov2013exploiting} suggest an alternate method for bilingual lexicon
-extraction, centered around two monolingual vector spaces $E$ (corresponding to the
-source language) and $F$ (corresponding to the target language). These vector spaces
-are constructed using a standard neural language model \cite{mikolov2013}.
-We see this as a promising alternative to the aforementioned bilingual vector space approach, as it allows for the construction of robust distributed representations whose quality is not dependent on manually selected seeds.
-The authors find
-that a linear model trained via stochastic gradient descent to map between these two
-vector spaces can effectively map source-language word representations to the
-target-language word vectors corresponding to their translations.
-
-We extend this work in attempting to model a \emph{translation function} $t : E \to F$.
-% SHORT
-% We capitalize on the fact that these vector spaces may be independently constructed, experimenting with different configurations for each space.
-In this paper we approximate the transformation $t$ between the two spaces using a deep neural
-network, trained to accept as input word vectors from the source language space
-$E$ and output word vectors in the target language space $F$.
+\newcite{mikolov2013exploiting} suggest an alternate method for bilingual
+lexicon extraction, centered around two monolingual vector spaces $E$
+(corresponding to the source language) and $F$ (corresponding to the target
+language). These vector spaces are constructed using a standard neural language
+model~\cite{mikolov2013}. This is a promising alternative to the aforementioned
+bilingual vector space approach, as it allows for the construction of robust
+distributed representations whose quality is not dependent on manually selected
+seeds. The authors find that a linear model trained via stochastic gradient
+descent to map between these two vector spaces can effectively find word-level
+translations from the source to the target language. We extend this work in
+attempting to approximate a \emph{translation function} $t : E \to F$ using a
+deep neural network, trained to accept as input word vectors from the source
+language space $E$ and output word vectors in the target language space $F$. We
+capitalize on the fact that these vector spaces may be independently
+constructed, experimenting with different configurations for each space.
 
@@
 
-Our translation algorithm makes use of two monolingual corpora: one consisting of text
-in the source language and the other containing text in the target language. We presume
-the corpora have the following properties:
-\subparagraph{Large and broad-domain.} With this property given, we can safely assume
-that for any common
-word which appears in the source-language corpus, we will also find its
-translation somewhere in the target-language corpus.
-If large corpora are unavailable, we can make use of smaller narrow-domain corpora, such as religious texts.
-\subparagraph{Non-parallel.} This means that our algorithm does not depend on any sort
-of alignment, lexical or topical. This relative flexibility allows us to demonstrate
-that the method can be applied in situations where parallel corpora are not available.
+Our translation algorithm makes use of two monolingual corpora: one consisting
+of text in the source language and the other containing text in the target
+language. We require only that the corpora have the following properties:
+
+\subparagraph{Large and broad-domain.} With this property given, we can safely
+assume that for any common word which appears in the source-language corpus, we
+will also find its translation somewhere in the target-language corpus. If large
+corpora are unavailable, we can make use of smaller narrow-domain corpora (e.g.,
+religious texts, famous novels).
+
+\subparagraph{Non-parallel.} This means that our algorithm does not depend on
+any sort of alignment, lexical or topical. This relative flexibility allows us
+to demonstrate that the method can be applied in situations where parallel
+corpora are not available.
 
@@
 In our experiments we use English and Spanish corpora sourced from Wikipedia.%
-\footnote{Available at \url{http://dumps.wikimedia.org/}.}
-We choose not to take advantage of the document alignments available in the Wikipedia corpus in order to show that our method is generalizable to other corpora.
+\footnote{Available at \url{http://dumps.wikimedia.org/}.} We choose not to take
+advantage of the document alignments available in the Wikipedia corpus in order
+to show that our method is generalizable to other corpora.
 
-In addition to the two monolingual corpora, we train on a minimal
-%, manually-created
-set of word translation pairs.
-% TODO I don't like "such as a swadesh list" --- Jon
-The specifics of these training pairs are described in \Cref{subsec:translation-nn}.
-
-%In addition to using the corpora, we also rely on a small seed set of several hundred English % -Spanish word pairs to train the translation function.
-% Include the seed sets as an appendix?
+In addition to the two monolingual corpora, we train on a minimal set of word
+translation pairs. The specifics of these training pairs are described in
+\Cref{subsec:translation-nn}.
 
@@
     }
-    \caption{The process of word translation, using a neural network to map between
-        two monolingual VSMs}
+    \caption{The process of word translation, using a neural network to map
+      between two monolingual VSMs}
     \label{fig:translation}
@@
 
-\Cref{fig:translation} illustrates the process of translating a single word from a source
-language to a target language (here English and Spanish, respectively) in this model. We
-begin with an input source-language word and retrieve a corresponding pre-computed word
-representation in the source VSM. We use this source-language word vector as an input to
-a neural network, which outputs a vector in the target-language VSM. We perform a simple
-nearest-neighbor search in the target-language VSM to find the closest existing word
-vector, and take its corresponding target-language word to be the translation of the
-original source-language word.
+\Cref{fig:translation} illustrates the process of translating a single word from
+a source language to a target language (here English and Spanish, respectively)
+in this model. We begin with an input source-language word and retrieve a
+corresponding pre-computed word representation in the source VSM\@. We use this
+source-language word vector as an input to a neural network, which outputs a
+vector in the target-language VSM\@. We perform a simple nearest-neighbor search
+in the target-language VSM to find the closest existing word vector, and take
+its corresponding target-language word to be the translation of the original
+source-language word.
 
@@
 
-For each of the two corpora introduced in \Cref{sec:data}, we train a continuous vector space model on 10-word contexts, using an implementation in the \textit{Gensim} software package of \texttt{word2vec}, a neural language model \cite{rehurek2010,mikolov2013}.
-This neural language model attempts to learn word representations which minimize error in the task of predicting a word's context given the word itself.
-The resulting word vectors are thus theoretically optimal (i.e., minimum-cost) representations of the meanings underlying their corresponding words.
-This is in contrast with the word representations built by a standard BLE approach, which define themselves in terms of manually constructed axes, and thus do not necessarily account for a maximal amount of variance in meaning among the words represented.
+For each of the two corpora introduced in \Cref{sec:data}, we train a continuous
+vector space model on 10-word contexts, using an implementation in the
+\textit{Gensim} software package of \texttt{word2vec}, a neural language
+model~\cite{rehurek2010,mikolov2013}. This neural language model attempts to
+learn word representations which minimize error in the task of predicting a
+word's context given the word itself. The resulting word vectors are thus
+theoretically optimal (i.e., minimum-cost) representations of the meanings
+underlying their corresponding words. This is in contrast with the word
+representations built by a standard BLE approach, which define themselves in
+terms of manually constructed axes, and thus do not necessarily account for a
+maximal amount of variance in meaning among the words represented. We experiment
+with different word vector dimensionalities, and detail our specific choices
+later in \Cref{subsec:test-data}.
 
-We experiment with different word vector dimensionalities, and detail our specific choices later in \Cref{subsec:test-data}.
-
-%A VSM is structured such that the relative position of a word encodes aspects of its meaning.
-%For example, similar words map into similar neighborhoods, and addition by the same vector can produce analogies (citation needed).\unnec
-% Ok, I suppose we don't need to give that much detail (especially for a 4 page paper).
-A brief mention, though, can help explain the intuition that some structure can be extrapolated by a neural network --- Arthur
-% Since independently-produced VSMs maintain similar patterns, this suggests we can build a translation function from a small number of seeds.
+While we focus on constructing a neural language model with \texttt{word2vec},
+we believe any distributional model of word meaning would suffice for this
+task~\cite{bengio2003,turian2010,collobert2011}.
 
@@
 
-The computation central to the process of translation is the transformation of a
-source-language word representation into a target-language word representation via
-a deep neural network. We learn this transformation from a small set of training pairs.
-%broad-domain training pairs, each mapping a single source language word to a single
-%translationally equivalent
-%target language word.
+The central computation is the transformation of a source-language word
+representation into a target-language word representation via a deep neural
+network. We learn this transformation from a small set of training pairs.
 
-% TODO added this.. what do you think? --- Jon
 The standard bilingual vector space approach detailed in the introduction would
-base its word representations on these training pairs. In contrast with this method,
-our model uses these translation pairs only in its final learning step, while training
-the neural network which models the translation function. For this reason, the model
-can benefit from effective word representations developed independently from the
-translation system itself. Furthermore, it can provide direct feedback through a training cost function
-on the quality of the translation pairs.
+base its word representations on these training pairs. In contrast with this
+method, our model uses these translation pairs only in its final learning step,
+while training the neural network which models the translation function. For
+this reason, the model can benefit from effective word representations developed
+independently from the translation system itself. Furthermore, it can provide
+direct feedback through a training cost function on the quality of the
+translation pairs.
 
-We investigate several different methods 
-of producing translation pairs.
-In line with previous
-approaches, we fill a set
-of translation pairs
-\texttt{Cog} with cognates shared between the two
-languages \cite{koehn2002}. We also build a small separate set
-\texttt{ConcNN}
-consisting of concrete nouns, and attempt to restrict the set to word pairs for which
-both the source- and target-language words have just one sense in their respective
-languages.
+We investigate several different methods of producing translation pairs. In line
+with previous approaches, we fill a set of translation pairs \texttt{Cog} with
+cognates shared between the two languages~\cite{koehn2002}. We also build a
+small separate set \texttt{ConcNN} consisting of concrete nouns, and attempt to
+restrict the set to word pairs for which both the source- and target-language
+words have just one sense in their respective languages.
 
-For each training pair of words $(e, f)$, we provide as a training example the input and output $(w_e, w_f)$, where $w_e$ and $w_f$ are word representations
-drawn from their respective VSMs constructed as described in \Cref{subsec:word-representations}.
+For each training pair of words $(e, f)$, we provide as a training example the
+input and output $(w_e, w_f)$, where $w_e$ and $w_f$ are word representations
+drawn from their respective VSMs constructed as described in
+\Cref{subsec:word-representations}.
 
-At inference time, for a given English word $e$, we provide as input to the neural network
-the corresponding word representation $w_e$. The output of the network, $w_f$, is some
-vector in the target-language vector space. We search for the nearest known neighbor of
-$w_f$, and treat the word corresponding to this nearest neighbor as the translation $f$
-of the input word $e$.
+At inference time, for a given English word $e$, we provide as input to the
+neural network the corresponding word representation $w_e$. The output of the
+network, $w_f$, is some vector in the target-language vector space. We search
+for the nearest known neighbor of $w_f$, and treat the word corresponding to
+this nearest neighbor as the translation $f$ of the input word $e$.
 
@@
 
-\newcite{mikolov2013exploiting} found that a linear model could learn a translation mapping
-between two monolingual vector spaces. We construct such a linear model for purposes of
-comparison with the neural network method detailed above. It utilizes
-the same basic idea as presented in \Cref{fig:translation}: it only differs in the type of
-transformation used to map between the two monolingual vector spaces. Whereas our approach uses
-a deep neural network, this linear model builds a linear transformation between the two spaces.
-Given source-language and target-language training data matrices $W_E, W_F$ (where the $i$th
-columns of the two matrices are the vector word representations corresponding to the $i$th
-training pair) we assert the existence of a translation matrix $A$, where $W_F = AW_E$.
+\newcite{mikolov2013exploiting} found that a linear model could learn a
+translation mapping between two monolingual vector spaces. We construct such a
+linear model for purposes of comparison with the neural network method detailed
+above. It utilizes the same basic idea as presented in \Cref{fig:translation}:
+it only differs in the type of transformation used to map between the two
+monolingual vector spaces. Whereas our approach uses a deep neural network, this
+linear model builds a linear transformation between the two spaces. Given
+source-language and target-language training data matrices $W_E, W_F$ (where the
+$i$th columns of the two matrices are the vector word representations
+corresponding to the $i$th training pair) we assert the existence of a
+translation matrix $A$, where $W_F = AW_E$.
 
-We compute the linear transformation by taking $A = W_F W_E^+$, where $W_E^+$ is the Moore--Penrose
-pseudoinverse of $W_E$.
-% SHORT
-% It is necessary to use the pseudoinverse of $W_E$, as the matrix is unlikely to be square and invertible.
-Then given any source-language word vector $w_e$, we calculate
-the transformed target-language word vector as $w_f = A w_e$.
+We compute the linear transformation by taking $A = W_F W_E^+$, where $W_E^+$ is
+the Moore-Penrose pseudoinverse of $W_E$. It is necessary to use the
+pseudoinverse of $W_E$, as the matrix is unlikely to be square and invertible.
+Then given any source-language word vector $w_e$, we calculate the transformed
+target-language word vector as $w_f = A w_e$.
 
@@
 \label{sec:evaluation}
-% TODO maybe for long paper: note that test sets mirror the style of the seed set. e.g.
-% model trained on concrete nouns is tested on unseen concrete nouns; model tested on
-% cognates is tested on unseen cognates
-We evaluate our translation model on an unseen set of English-Spanish word pairs.
-Since our model builds a ranked list of nearest-neighbor translation candidates, we use the mean reciprocal rank (MRR) of the correct translation within the list:
+We evaluate our translation model on an unseen set of English-Spanish word
+pairs. These test sets are customized to match the algorithm's training
+environment. For example, a translation algorithm trained on concrete nouns from
+the training set \texttt{ConcNN} will be tested on other unseen concrete nouns.
+Since our model builds a ranked list of nearest-neighbor translation candidates,
+we can calculate the mean reciprocal rank (MRR) of the correct translation
+within the list:
 \begin{equation}
@@
 \end{equation}
-where $r_j$ is the rank of the correct translation for pair $j$ out of $n$.
-%For example, if every correct translation is ranked second highest, $\textit{MRR} = \frac{1}{2}$. % leaving out for space considerations
+where $r_j$ is the rank of the correct translation for pair $j$ out of $n$. For
+example, a mean reciprocal rank of 1 would indicate that every translation
+output ranked the correct translation in the first position. A mean reciprocal
+rank of $\frac{1}{2}$ would indicate that the average translation output placed
+the correct translation as the second most likely in a ranked list.
 
@@
 
-We establish a lower bound for translation performance with an algorithm that associates
-source- and target-language words by their relative corpus frequencies. Given an input
-source-language word, we compute its corpus frequency percentile. We then predict
-translations in the target language by returning the words with the closest percentile
-frequencies in the target language corpus.
+We establish a lower bound for translation performance with an algorithm that
+associates source- and target-language words by their relative corpus
+frequencies. Given an input source-language word, we compute its corpus
+frequency percentile. We then predict translations in the target language by
+returning the words with the closest percentile frequencies in the target
+language corpus.
 
@@
 
-We establish three separate test environments, each defined by a choice of training data,
-input corpus format and VSM design. The vector space dimensionalities given below
-were derived through trial and error.
-% SHORT We find that different VSM dimensionalities are optimal for different corpus inputs.
-Overall, we see that performance varies significantly based on these test environment properties.
-The environments are as follows:
+We establish three separate test environments, each defined by a choice of
+training data, input corpus format, and VSM design. The vector space
+dimensionalities given below were derived through trial and error. We find that
+different VSM dimensionalities are optimal for different corpus inputs. Overall,
+we see that performance varies significantly based on these test environment
+properties. The environments are as follows:
 \begin{description}
-\item[\texttt{Cog}:] Minimally preprocessed corpora, \texttt{Cog} training data, 500-dimensional word vectors % TODO long paper: 5000/750 min word threshold
-% Perhaps note that we don't do any Levenshtein mumbo-jumbo? --- Arthur
-\item[\texttt{ConcNN}:] Minimally preprocessed corpora, \texttt{ConcNN} training data, 500-dimensional word vectors % TODO long paper: 10000/1500 min word threshold
-\item[\texttt{CogLem}:] Lemmatized Spanish corpus, \texttt{Cog} training data, 1000-dimensional word vectors % TODO long paper: 2500/375 min word threshold
+\item[\texttt{Cog}:] Minimally preprocessed corpora, \texttt{Cog} training data,
+  500-dimensional word vectors % TODO long paper: 5000/750 min word threshold
+\item[\texttt{ConcNN}:] Minimally preprocessed corpora, \texttt{ConcNN} training
+  data, 500-dimensional word
+  vectors % TODO long paper: 10000/1500 min word threshold
+\item[\texttt{CogLem}:] Lemmatized Spanish corpus, \texttt{Cog} training data,
+  1000-dimensional word vectors % TODO long paper: 2500/375 min word threshold
 \end{description}
@@
 
-Our evaluation metric looks for a single possible translation of a test input word in
-ranked output. As we explain in \Cref{subsec:error-analysis}, this may needlessly penalize
-the algorithm.
-% SHORT , as it often suggests reasonable translations earlier in the ranked output which happen to not match our strict search requirements.
-For example, our algorithm is
-tested on a cognate pair $(\textit{concept}, \textit{concepto})$, and is penalized for
-suggesting the translation \textit{idea}, a suitable synonym of the desired target word.
-A fourth test run \bt{CogFix} considers the algorithm's output on the \texttt{Cog} dataset,
-and accepts such sensible alternate translations. % SHORT (rather than stipulating in this case that all source-language cognates translate to their corresponding target-language cognates).
+Our evaluation metric looks for a single possible translation of a test input
+word in ranked output. As we explain in \Cref{subsec:error-analysis}, this may
+needlessly penalize the algorithm, as it often suggests reasonable translations
+earlier in the ranked output which happen to not match our strict search
+requirements. For example, our algorithm is tested on a cognate pair
+$(\textit{concept}, \textit{concepto})$, and is penalized for suggesting the
+translation \textit{idea}, a suitable synonym of the desired target word. A
+fourth test run \bt{CogFix} considers the algorithm's output on the \texttt{Cog}
+dataset, and accepts such sensible alternate translations (rather than
+stipulating in this case that all source-language cognates translate to their
+corresponding target-language cognates).
 
@@
 Mean reciprocal rank values for each test environment and a number of training
-conditions are in \Cref{tbl:results}.
-The linear transformation model based on \newcite{mikolov2013exploiting} was very competitive with the neural network model.
-This suggests that many---but not all---of the geometric relationships in the target language vector space can be modeled through a simple linear transform of the geometric structures in the source language space.
+conditions are in \Cref{tbl:results}. The linear transformation model based on
+\newcite{mikolov2013exploiting} was very competitive with the neural network
+model. This suggests that many---but not all---of the geometric relationships in
+the target language vector space can be modeled through a simple linear
+transform of the geometric structures in the source language space.
 
-The best absolute performance is achieved by a neural network with 250 hidden-layer
-neurons in the \bt{CogFix} training environment. Interestingly, the hidden layer of this model is of
-lower dimensionality than the test input and output word vectors (each 500 dimensions).
-% TODO repetitive "This suggests" --- Jon
-% Mention that the Linear model may be overfitting, if compressing first leads to better results? --- Arthur
-This suggests that the model learns some useful compressed intermediate representation of
-the input source-language word in the translation training process.
-% SHORTCAND
-Its MRR shows us that on average the algorithm suggests an acceptable translation in the second or third position of the ranked list of possible translations.
+The best absolute performance is achieved by a neural network with 250
+hidden-layer neurons in the \bt{CogFix} training environment. Interestingly, the
+hidden layer of this model is of lower dimensionality than the test input and
+output word vectors (each 500 dimensions). This suggests that the model learns
+some useful compressed intermediate representation of the input source-language
+word in the translation training process. Its MRR shows us that on average the
+algorithm suggests an acceptable translation in the second or third position of
+the ranked list of possible translations.
 
@@
 algorithm's output in \Cref{tbl:errors}, and elaborate on the classes of errors below:
-\subparagraph{Synonyms.} With many of the "errors" of this form, the algorithm provided an acceptable alternate translation that we simply did
-not expect. These cases are treated as successes in the \bt{CogFix} training environment, as described
-in \Cref{subsec:results}.
-\subparagraph{Near misses.} In these cases the algorithm mapped into a precise neighborhood
-of related words, but picked the wrong element from the set. % SHORT
-% For example, the output word \textit{chocolate} is strongly related to the expected word \textit{vainilla}.
-% TODO long paper: future work might examine the actual distance values and "give up" if the top few words are all very close
-\subparagraph{Right neighborhood, wrong spot.} We occasionally mapped into a correct larger
-area of the target language vector space, but still made a significant semantic error.
-% SHORT
-% With the input word \textit{election}, for example, the algorithm predicts as translations target-language words related to political campaigns.
+
+\subparagraph{Synonyms.} With many of the "errors" of this form, the algorithm
+provided an acceptable alternate translation that we simply did not expect.
+These cases are treated as successes in the \bt{CogFix} training environment, as
+described in \Cref{subsec:results}.
+
+\subparagraph{Near misses.} In these cases the algorithm mapped into a precise
+neighborhood of related words, but picked the wrong element from the set. For
+example, the output word \textit{chocolate} is strongly related to the expected
+word \textit{vainilla}. % TODO long paper: future work might examine
+% the actual distance values and "give up" if the top few words are all very
+% close
+
+\subparagraph{Right neighborhood, wrong spot.} We occasionally mapped into a
+correct larger area of the target language vector space, but still made a
+significant semantic error. With the input word \textit{election}, for example,
+the algorithm predicts as translations target-language words related to
+political campaigns.
+
 \subparagraph{Alternate senses.} In several cases on the \texttt{Cog} training
-set and its derivatives, the algorithm learned to predict translations of the input words which
-corresponded to valid alternate interpretations of the source language word.
-% SHORT
-% For example, it interprets the English \textit{division} in the sense of a military unit (rather than in the mathematical sense as we expected).
-Note that these cases were not treated as
-successes in the \bt{CogFix} training environment. We hope to eliminate this class of errors
-in future work by using multi-prototype VSMs (see \Cref{sec:conclusion}).
+set and its derivatives, the algorithm learned to predict translations of the
+input words which corresponded to valid alternate interpretations of the source
+language word. For example, it interprets the English \textit{division} in the
+sense of a military unit (rather than in the mathematical sense as we expected).
+These cases were not treated as successes in the \bt{CogFix} training
+environment. We hope to eliminate this class of errors in future work by using
+multi-prototype VSMs (see \Cref{sec:conclusion}).
 
@@
 
-In this paper, we presented an algorithm which learns word translations from unlabeled,
-non-parallel corpora by building a mapping between two independently constructed
-monolingual vector spaces. This algorithm can be used to perform bilingual lexicon
-extraction in scenarios where the source and target language in question do not benefit
-from widely available parallel corpora. We demonstrated encouraging translation
-performance with large, broad-domain corpora as text sources.
+In this paper, we presented an algorithm which learns word translations from
+unlabeled, non-parallel corpora by building a mapping between two independently
+constructed monolingual vector spaces. This algorithm can be used to perform
+bilingual lexicon extraction in scenarios where the source and target language
+in question do not benefit from widely available parallel corpora. We
+demonstrated encouraging translation performance with large, broad-domain
+corpora as text sources.
 
-For simplicity, our current VSM design associates a single word vector with each word
-in the source or target language corpus. However, training error could be greatly
-reduced by allowing polysemous words to be associated with multiple word vectors
-\cite{reisinger2010,huang2012}. We believe such an extension will remove much of the remaining
-ambiguity in the input vector spaces.
+For simplicity, our current VSM design associates a single word vector with each
+word in the source or target language corpus. However, training error could be
+greatly reduced by allowing polysemous words to be associated with multiple word
+vectors~\cite{reisinger2010,huang2012}. We believe such an extension will remove
+much of the remaining ambiguity in the input vector spaces.
 

### 2014-06-07 paper/main.tex
Add note in future work about using actual distance measure values

@@
 example, the output word \textit{chocolate} is strongly related to the expected
-word \textit{vainilla}. % TODO long paper: future work might examine
-% the actual distance values and "give up" if the top few words are all very
-% close
+word \textit{vainilla}. We describe a possible solution to this type of error in
+\Cref{sec:conclusion}.
 
@@
 
+As mentioned in \Cref{subsec:error-analysis}, a substantial amount of
+translation errors encountered during testing were due to ``near misses'': an
+incorrect selection of a target-language word from a correctly-chosen precise
+semantic neighborhood. Future work might examine the actual distance values
+between the algorithm's target-language vector output and the target-language
+word vectors which we compare to, and make more intelligent translation choices
+(perhaps refusing to provide exactly one output value) when the translation
+seems too semantically ambiguous. % TODO unclear?
+
 % However, as we note that many of our errors are caused by alternate sense interpretations, our approach could be further improved through disambiguation and multiple vector representations per word.

### 2014-06-07 paper/main.tex
Minor formatting changes

@@
 corresponding to the $i$th training pair) we assert the existence of a
-translation matrix $A$, where $W_F = AW_E$.
+translation matrix $A$, where
+\begin{equation}
+  W_F = AW_E.
+\end{equation}
 
@@
 Then given any source-language word vector $w_e$, we calculate the transformed
-target-language word vector as $w_f = A w_e$.
+target-language word vector as
+\begin{equation}
+  w_f = A w_e.
+\end{equation}
 
@@
   \centering
-    \small
+  \resizebox{\linewidth}{!}{
     \begin{tabular}[t]{l lll}
@@
     \end{tabular}
+  }
   \caption{Examples of common error patterns observed in the algorithm's output for

### 2014-06-08 paper/main.tex
Add maybe-useful extra future work

@@
 
+Lastly, future work might acknowledge the questionable assumption that there
+exists a \emph{function} between source- and target-language words: that is,
+that there is one target-language output for each source-language input. This
+might be overcome by explicitly working to map between VSM clusters (perhaps
+filtered by part of speech) rather than precise words. % TODO worth including?
+
 % However, as we note that many of our errors are caused by alternate sense interpretations, our approach could be further improved through disambiguation and multiple vector representations per word.

### 2014-07-15 model/runner.py
Unicode-friendly translation logging + log index of correct translation

@@ -34,11 +34,17 @@ def get_translations(source_word, threshold=10000):
 
 
 def log_translations(source_word, target_word, translations):
-    targets = [u'\033[1m**{}**\033[0m'.format(target)
-               if target == target_word else target
-               for target in translations]
-    logging.info(u'Translations of {}: {}'
-                 .format(source_word, u' '.join(targets)))
+    targets = [u'\033[1m**{}**\033[0m'.format(target.decode('utf-8'))
+               if target == target_word else target.decode('utf-8')
+               for target in translations[:100]]
+
+    try:
+        pos = translations.index(target_word) + 1
+    except ValueError:
+        pos = 0
+
+    logging.info(u'Translations of {} ({}): {}'
+                 .format(source_word, pos, u' '.join(targets)))
 
 
 def score(source_word, expected_target_word, translations):

### 2014-07-15 paper/main.tex
added Arthur contributions

@@
 
+\section{Contributions}
+\label{sec:contributions}
+
+Jon developed the translation framework in Python, building off several existing tools (notably Word2Vec, Gensim, and Pylearn).
+He trained a suite of models and evaluated them on manually sourced translation seeds.
+
+Arthur also applied this method to translate from the King James Bible
+to a modern English translation. The results there were not so
+interesting as they mostly mapped the same words onto themselves. He
+also wrote a tool to map the vector spaces in t-SNE, but this is still
+a work in progress.
+
 \bibliographystyle{acl}
